---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# njoagleod

<!-- badges: start -->
[![R-CMD-check](https://github.com/tor-gu/njoagleod/workflows/R-CMD-check/badge.svg)](https://github.com/tor-gu/njoagleod/actions)
<!-- badges: end -->

This is a cleaned-up version of the NJ OAG Law Enforcement Officer Diversity 
  database available from https://www.njoag.gov/policerecruiting/

## Dataset Overview
This dataset contains two tables.

* ```agency``` Has one row for each law enforcement agency, for each year
* ```officer``` Has one row for each officer, for each year.

Currently 2021 is the only year represented.

## Examples
The ```agency``` table looks like this:
```{r}
library(njoagleod)
library(tidyverse)
agency
```
The columns ```year```, ```agency_county``` and ```agency_name``` form the primary key.  

The ```officer``` table looks like this.
```{r}
officer
```
The columns ```year```, ```agency_county``` and ```agency_name``` act as a foreign
key to the ```agency``` table.

The ```agency_name``` and ```agency_county``` columns are consistent with the columns
of the same name in the associated [```njoaguof```](https://github.com/tor-gu/njoaguof) package so that they may be easily used together.  

For example, here are the agencies with the highest number of use-of-force incidents per officer.
```{r}
officer %>% 
  group_by(year, agency_county, agency_name) %>%
  summarize(officer_count = n(), .groups = "drop") %>%
  left_join(njoaguof::incident, by=c("agency_county", "agency_name")) %>%
  filter(lubridate::year(incident_date_1)==year) %>%
  group_by(year, agency_county, agency_name, officer_count) %>%
  summarize(incident_count = n(), .groups="drop") %>%
  mutate(incidents_per_officer = incident_count / officer_count) %>%
  select(-agency_county) %>%
  arrange(desc(incidents_per_officer))
```

## Notes

### Duplicates entries removed
In the source data, The Delaware River and Bay Authority Police Department and its officers appear twice, once associated to Salem County and once associated to Cape May County. 

In this package, we have removed the duplicate entries and assigned the county to `NA`.  As a result, the source data has ```30617``` officers across ```530``` agencies, while this package contains ```30565``` officers and ```529``` agencies.

### Chester PD `agency_type` changed to `"Multiple Municipality"`
In the source data, Chester PD is listed as a municipal agency, but it is a 
multiple municipality agency. In this package, the agency type has been changed to "Multiple Municipality".

Source: [Former site of the Chester Borough Police](https://chesterborough.org/departments/chester-borough-police/)

## Installation

You can install the latest version of njoaguof from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("tor-gu/njoagleod")
```

