---
title: "NJ OAG Law Enforcement Officer Diversity data cleaning workbook"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NJ OAG Law Enforcement Officer Diversity data cleaning workbook}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(magrittr)
library(stringr)
library(tidyr)
library(njoaguof)
```

# Read the raw data
```{r}
agency_2021_csv <- "2021_NJOAGLEOD_Agency_Information.csv"
officer_2021_csv <- "2021_NJOAGLEOD_Officer_Information.csv"

agency_2021 <- system.file("extdata", agency_2021_csv, package="njoagleod") %>%
  readr::read_csv()
officer_2021 <- system.file("extdata", officer_2021_csv, package="njoagleod") %>%
  readr::read_csv()
```

# Clean up the `agency` table
Rename the columns and add a year column.
```{r}
agency <- agency_2021 %>%
  rename(agency_county=County,
         agency_name=Agency,
         agency_type=`Agency Type`,
         hiring_governed_by_civil_service=`Hiring governed by Civil Service?`) %>%
  mutate(year=2021) %>% 
  relocate(year)
```

## `agency_county`
We want to use the same counties as the US census (and package `njoaguof`).
In the source data, non-county agencies use "Statewide Or Regional LEA". as the
county. We will map this to `NA`.  We put this in a function, because we will
do this twice -- once for the agency table, and once for the officer table.
```{r}
fix_agency_county <- function(tbl) {
  county_levels <- levels(njoaguof::incident$agency_county)
  stopifnot(
    "Statewide Or Regional Lea County" ==
      tbl %>% pull(agency_county) %>% str_to_title() %>%
      paste0(" County") %>% setdiff(county_levels)
  )
  tbl %>%
    mutate(agency_county=str_to_title(agency_county)) %>%
    mutate(agency_county=paste0(agency_county, " County")) %>%
    mutate(agency_county=factor(agency_county, county_levels))
}

agency <- agency %>% fix_agency_county()
```
## `agency_name`
We want to use the same agency names as the `njoaguof` package, so we will
do some cleanup here.  Because we need to do this for both the `agency` and
`officer` tables, we put it in a function.

```{r}
fix_agency_name <- function(tbl) {
  tbl %>%
    mutate(
      agency_name = case_when(
        agency_name == "Sussex Co Sheriff'S Dept" ~ "Sussex Co Sheriffs Dept",
        agency_name == "Cumberland Co Sheriff'S Dept" ~ "Cumberland Co Sheriffs Dept",
        agency_name == "Hunterdon Co Sheriff'S Dept" ~ "Hunterdon Co Sheriffs Dept",
        agency_name == "Burlington County Sheriffs Off" ~ "Burlington Co Sheriffs Office",
        agency_name == "Mercer Co Prosecutor Offfice" ~ "Mercer Co Prosecutors Office",
        agency_name == "Hudson Co Prosecutor Office" ~ "Hudson Co Prosecutors Office",
        agency_name == "Somerset County Sheriff's Office" ~ "Somerset Co Sheriffs Office",
        agency_name == "Totowa Boro  PD" ~ "Totowa Boro PD",
        agency_name == "New Jersey Institute Of Technol" ~ "New Jersey Institute Of Technology",
        agency_name == "NJ Div Criminal Justice - Trenton" ~ "NJ Division of Criminal Justice",
        TRUE ~ agency_name
      )
    )
}

agency <- agency %>% fix_agency_name()
```
## `agency_type`
There are four values we expect here. We encode this as a factor and
check that there are no rows which don't match.
```{r}
agency_type_levels <- c("State-wide", "County", "Municipal", 
                        "Multiple Municipalities", "Not Provided")
agency <- agency %>%
  mutate(agency_type = factor(agency_type, agency_type_levels))

stopifnot(0 == agency %>% filter(is.na(agency_type)) %>% nrow())
```

## `hiring_governed_by_civil_service`
This is a yes/no field (with some inconsistency in captialization) which we convert to a boolean. `"Not provided"` and other values are mapped to `NA`. 
```{r}
agency <- agency %>% 
  mutate(hiring_governed_by_civil_service = 
           str_to_lower(hiring_governed_by_civil_service)) %>%
  mutate(hiring_governed_by_civil_service=case_when(
    hiring_governed_by_civil_service == "yes" ~ TRUE,
    hiring_governed_by_civil_service == "no" ~ FALSE,
    TRUE ~ NA
  )) 
```

# Clean up the `officer` table
Rename the columns and add a year column.
```{r}
officer <- officer_2021 %>%
  rename(agency_county=County,
         agency_name=Agency,
         officer_age=Age,
         officer_race=`Race/Ethnicity`,
         officer_gender=Gender2
         ) %>%
  mutate(year=2021) %>% 
  relocate(year)
```

## `agency_county` and `agency_name`
We reuse the functions created to the `agency` table.
```{r}
officer <- officer %>% 
  fix_agency_county() %>% 
  fix_agency_name()
```

## `officer_age`
No change.

## `officer_race`
We use the same levels as the `njoaguof` package, mapping `"Not provided"` to 
`NA`.  We will do a check to make sure that we are not missing any values when
we do this mapping.

```{r}
# Save this value for the check at the end
officer_race_not_provided <- officer %>% 
  filter(officer_race=="Not provided") %>%
  nrow()

# Now do the mapping
race_levels <- levels(incident$officer_race)
officer <- officer %>% 
  mutate(officer_race = str_to_title(officer_race)) %>%
  mutate(officer_race = case_when(
    officer_race == "American Indian Or Alaska Native" ~ "American Indian",
    officer_race == "Black Or African American" ~ "Black",
    officer_race == "Native Hawaiian Or Other Pacific Islander" ~ "Native Hawaiian or other Pacific Islander",
    officer_race == "Not Provided" ~ NA_character_,
    officer_race == "Two Or More Races" ~ "Two or more races",
    TRUE ~ officer_race
  )) %>% 
  mutate(officer_race=factor(officer_race, levels=race_levels))

# Check that we did not introduce any more `NA`s than we meant to.
stopifnot(
  officer_race_not_provided == 
    officer %>% filter(is.na(officer_race)) %>% nrow()
)
```

## `officer_gender`
We will use this same levels as the `njoaguof` package, mapping `"No Provided"` to `NA`. We will do a check to make sure that we are not missing any values when we do this mapping.
```{r}
# Save this value for the check at the end
officer_gender_not_provided <- officer %>% 
  filter(officer_gender=="Not Provided") %>%
  nrow()

# Now do the mapping
gender_levels <- levels(njoaguof::incident$officer_gender)
officer <- officer %>% 
  mutate(officer_gender=case_when(
    officer_gender == "Not Provided" ~ NA_character_,
    officer_gender == "X Or Non-Binary" ~ "Gender Non-Conforming/X",
    TRUE ~ officer_gender
  )) %>%
  mutate(officer_gender=factor(officer_gender, levels=gender_levels))

# Check that we did not introduce any more `NA`s than we meant to.
stopifnot(
  officer_gender_not_provided == 
    officer %>% filter(is.na(officer_gender)) %>% nrow()
)
```
