---
title: "NJ OAG Law Enforcement Officer Diversity data cleaning workbook"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NJ OAG Law Enforcement Officer Diversity data cleaning workbook}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(magrittr)
library(stringr)
library(tidyr)
library(njoaguof)
```

# Read the raw data
```{r}
agency_2021_csv <- "2021_NJOAGLEOD_Agency_Information.csv"
officer_2021_csv <- "2021_NJOAGLEOD_Officer_Information.csv"

agency_2021 <- system.file("extdata", agency_2021_csv, package="njoagleod") %>%
  readr::read_csv()
officer_2021 <- system.file("extdata", officer_2021_csv, package="njoagleod") %>%
  readr::read_csv()
```

# Clean up the `agency` table
Rename the columns and add a year column.
```{r}
agency <- agency_2021 %>%
  rename(agency_county=County,
         agency_name=Agency,
         agency_type=`Agency Type`,
         hiring_governed_by_civil_service=`Hiring governed by Civil Service?`) %>%
  mutate(year=2021) %>% 
  relocate(year)
```

## `agency_county`
We want to use the same counties as the US census (and package `njoaguof`).
In the source data, non-county agencies use "Statewide Or Regional LEA". as the
county. We will map this to `NA`.  We put this in a function, because we will
do this twice -- once for the agency table, and once for the officer table.
```{r}
fix_agency_county <- function(tbl) {
  county_levels <- levels(njoaguof::incident$agency_county)
  stopifnot(
    "Statewide Or Regional Lea County" ==
      tbl %>% pull(agency_county) %>% str_to_title() %>%
      paste0(" County") %>% setdiff(county_levels)
  )
  tbl %>%
    mutate(agency_county=str_to_title(agency_county)) %>%
    mutate(agency_county=paste0(agency_county, " County")) %>%
    mutate(agency_county=factor(agency_county, county_levels))
}

agency <- agency %>% fix_agency_county()
```
## `agency_name`
We want to use the same agency names as the `njoaguof` package, so we will
do some cleanup here.  Because we need to do this for both the `agency` and
`officer` tables, we put it in a function.

```{r}
fix_agency_name <- function(tbl) {
  tbl %>%
    mutate(
      agency_name = case_when(
        agency_name == "Sussex Co Sheriff'S Dept" ~ "Sussex Co Sheriffs Dept",
        agency_name == "Cumberland Co Sheriff'S Dept" ~ "Cumberland Co Sheriffs Dept",
        agency_name == "Hunterdon Co Sheriff'S Dept" ~ "Hunterdon Co Sheriffs Dept",
        agency_name == "Burlington County Sheriffs Off" ~ "Burlington Co Sheriffs Office",
        agency_name == "Mercer Co Prosecutor Offfice" ~ "Mercer Co Prosecutors Office",
        agency_name == "Hudson Co Prosecutor Office" ~ "Hudson Co Prosecutors Office",
        agency_name == "Somerset County Sheriff's Office" ~ "Somerset Co Sheriffs Office",
        agency_name == "Totowa Boro  PD" ~ "Totowa Boro PD",
        agency_name == "New Jersey Institute Of Technol" ~ "New Jersey Institute Of Technology",
        agency_name == "NJ Div Criminal Justice - Trenton" ~ "NJ Division of Criminal Justice",
        agency_name == "Toms River Township" ~ "Toms River Twp PD",
        agency_name == "Hardyston Twp Police Dept" ~ "Hardyston Twp PD",
        agency_name == "Linden Police Department" ~ "Linden PD",
        TRUE ~ agency_name
      )
    )
}

agency <- agency %>% fix_agency_name()
```
## `agency_type`
There are four values we expect here. We encode this as a factor and
check that there are no rows which don't match.
```{r}
agency_type_levels <- c("State-wide", "County", "Municipal", 
                        "Multiple Municipalities", "Not Provided")
agency <- agency %>%
  mutate(agency_type = factor(agency_type, agency_type_levels))

stopifnot(0 == agency %>% filter(is.na(agency_type)) %>% nrow())
```

The Chester Police Department is incorrectly classified as municipal. It is a
multi-municipality agency covering Chester Borough and Chester Township. We make
that correction here.
```{r}
agency <- agency %>%
  mutate(agency_type = if_else(
    agency_name == "Chester PD", 
    factor("Multiple Municipalities", agency_type_levels), agency_type))
```


## `hiring_governed_by_civil_service`
This is a yes/no field (with some inconsistency in captialization) which we convert to a boolean. `"Not provided"` and other values are mapped to `NA`. 
```{r}
agency <- agency %>% 
  mutate(hiring_governed_by_civil_service = 
           str_to_lower(hiring_governed_by_civil_service)) %>%
  mutate(hiring_governed_by_civil_service=case_when(
    hiring_governed_by_civil_service == "yes" ~ TRUE,
    hiring_governed_by_civil_service == "no" ~ FALSE,
    TRUE ~ NA
  )) 
```

## For municipal agencies, add `agency_municipality`, if possible
Most municipal agency names are of the form `<town-name> PD`, but the `<town-name` does not match exactly with the name in the US census. We want to add the US Census name if it can be determined from the agency name.

First, lets verify that the the rule that municipal agency names end in "PD" has only three exceptions:
```{r}
stopifnot(
  setequal(
    agency %>% 
      filter(agency_type == "Municipal") %>%
      filter(!str_detect(agency_name, "PD$")) %>%
      pull(agency_name),
    c("Cherry Hill Campus Police", 
      "Stevens Institute Of Technology", 
      "Princeton University Police")
  )
)
```
Next, lets initialize a lookup table which we will fill up incrementally.  We include only municipal agencies that end in "PD". We temporarily add the column `agency_no_pd`, which is just the agency name minus the "PD" extension.
```{r}
municipality_lookup <- agency %>% 
  filter(agency_type == "Municipal") %>%
  filter(str_detect(agency_name, " PD$")) %>%
  mutate(agency_no_pd = str_remove(agency_name, " PD$")) %>%
  mutate(municipality = NA_character_) %>%
  select(agency_county, agency_name, agency_no_pd, municipality)
municipality_lookup
```
The following function will take a partially filled in `municipality_lookup` table with an added column -- `test_municipality` -- and compare it to the `municipality` table. Any matching row will be updated.

```{r}
add_test_municipalities <- function(municipality_lookup) {
  municipality_lookup %>% 
    filter(is.na(municipality)) %>%
    select(agency_name, county=agency_county, municipality=test_municipality) %>%
    inner_join(njoagleod::municipality, by=c("county", "municipality")) %>%
    right_join(municipality_lookup, by=c("agency_name", "county"="agency_county")) %>%
    mutate(municipality = if_else(
      is.na(municipality.y), 
      municipality.x, 
      municipality.y)) %>%
    select(agency_county=county, agency_name, agency_no_pd, municipality)
}
```
Now we try variations on the `agency_no_pd` column to fill out the `municipality_lookup` table.  We leave in the commented out `filter(is.na(municipality))` lines to emphasize the iterative nature of this step.
```{r}
municipality_lookup <- municipality_lookup %>% 
  #filter(is.na(municipality))
  mutate(test_municipality = paste0(agency_no_pd, " city")) %>%
  add_test_municipalities() %>% 
  #filter(is.na(municipality))
  mutate(test_municipality = paste0(agency_no_pd, " town")) %>%
  add_test_municipalities() %>% 
  #filter(is.na(municipality))
  mutate(test_municipality = paste0(agency_no_pd, " village")) %>%
  add_test_municipalities() %>% 
  #filter(is.na(municipality))
  mutate(test_municipality = paste0(agency_no_pd, " borough")) %>%
  add_test_municipalities() %>% 
  #filter(is.na(municipality))
  mutate(test_municipality = paste0(agency_no_pd, " township")) %>%
  add_test_municipalities() %>% 
  #filter(is.na(municipality))
  mutate(test_municipality = str_replace(agency_no_pd, "City", "city")) %>%
  add_test_municipalities() %>% 
  #filter(is.na(municipality))
  mutate(test_municipality = str_replace(agency_no_pd, "Twp", "township")) %>%
  add_test_municipalities() %>% 
  #filter(is.na(municipality))
  mutate(test_municipality = str_replace(agency_no_pd, "Boro", "borough")) %>%
  add_test_municipalities() %>% 
  #filter(is.na(municipality))
  mutate(test_municipality = str_replace(agency_no_pd, "Town", "town")) %>%
  add_test_municipalities() %>% 
  #filter(is.na(municipality))
  mutate(test_municipality = str_replace(agency_no_pd, "Village", "village")) %>%
  add_test_municipalities() %>% 
  #filter(is.na(municipality))
  mutate(test_municipality = paste0(agency_no_pd, " City city")) %>%
  add_test_municipalities()
```
For most of the remaining municipalities, we can set the `municipality` manually.
```{r}
municipality_lookup <- municipality_lookup %>% 
  mutate(municipality = case_when(
    agency_name == "Mt. Ephraim PD" ~ "Mount Ephraim borough",
    agency_name == "Orange PD" ~ "City of Orange township",
    agency_name == "South Orange PD" ~ "South Orange Village township",
    agency_name == "Avon-By-The-Sea PD" ~ "Avon-by-the-Sea borough",
    agency_name == "Bayhead Boro PD" ~ "Bay Head borough",
    agency_name == "Oceangate Boro PD" ~ "Ocean Gate borough",
    agency_name == "Peapack-Gladstone PD" ~ "Peapack and Gladstone borough",
    TRUE ~ municipality
  ))
```
At this point, there are two known exceptional agencies left. Let us remove them and then verify that the table is complete, and that every municipality in the lookup table is found in the `municipality` table.  We also don't need the `agency_no_pd` column anymore, so we will remove it.
```{r}
municipality_lookup <- municipality_lookup %>% 
  filter(!(agency_name %in%
             c("Monmouth University PD",
               "Washington Park PD"))) %>%
  select(-agency_no_pd)

# Check there are no mor NAs
stopifnot(0 ==
            municipality_lookup %>% filter(is.na(municipality)) %>% nrow())

# Check that every municipality is in the lookup table
stopifnot(0 ==
            municipality_lookup %>%
            anti_join(njoagleod::municipality, by = c("agency_county" = "county", "municipality")) %>%
            nrow())
```
Finally, we can use the lookup table to add a `municipality` column to the agency table.
```{r}
agency <- agency %>% 
  left_join(municipality_lookup, by=c("agency_county", "agency_name"))
```

# Clean up the `officer` table
Rename the columns and add a year column.
```{r}
officer <- officer_2021 %>%
  rename(agency_county=County,
         agency_name=Agency,
         officer_age=Age,
         officer_race=`Race/Ethnicity`,
         officer_gender=Gender
         ) %>%
  mutate(year=2021) %>% 
  relocate(year)
```

## `agency_county` and `agency_name`
We reuse the functions created to the `agency` table.
```{r}
officer <- officer %>% 
  fix_agency_county() %>% 
  fix_agency_name()
```

## `officer_age`
No change.

## `officer_race`
We use the same levels as the `njoaguof` package, mapping `"Not provided"` to 
`NA`.  We will do a check to make sure that we are not missing any values when
we do this mapping.

```{r}
# Save this value for the check at the end
officer_race_not_provided <- officer %>% 
  filter(officer_race=="Not provided") %>%
  nrow()

# Now do the mapping
race_levels <- levels(incident$officer_race)
officer <- officer %>% 
  mutate(officer_race = str_to_title(officer_race)) %>%
  mutate(officer_race = case_when(
    officer_race == "American Indian Or Alaska Native" ~ "American Indian",
    officer_race == "Black Or African American" ~ "Black",
    officer_race == "Native Hawaiian Or Other Pacific Islander" ~ "Native Hawaiian or other Pacific Islander",
    officer_race == "Not Provided" ~ NA_character_,
    officer_race == "Two Or More Races" ~ "Two or more races",
    TRUE ~ officer_race
  )) %>% 
  mutate(officer_race=factor(officer_race, levels=race_levels))

# Check that we did not introduce any more `NA`s than we meant to.
stopifnot(
  officer_race_not_provided == 
    officer %>% filter(is.na(officer_race)) %>% nrow()
)
```

## `officer_gender`
We will use this same levels as the `njoaguof` package, mapping `"No Provided"` to `NA`. We will do a check to make sure that we are not missing any values when we do this mapping.
```{r}
# Save this value for the check at the end
officer_gender_not_provided <- officer %>% 
  filter(officer_gender=="Not Provided") %>%
  nrow()

# Now do the mapping
gender_levels <- levels(njoaguof::incident$officer_gender)
officer <- officer %>% 
  mutate(officer_gender=case_when(
    officer_gender == "Not Provided" ~ NA_character_,
    officer_gender == "X Or Non-Binary" ~ "Gender Non-Conforming/X",
    TRUE ~ officer_gender
  )) %>%
  mutate(officer_gender=factor(officer_gender, levels=gender_levels))

# Check that we did not introduce any more `NA`s than we meant to.
stopifnot(
  officer_gender_not_provided == 
    officer %>% filter(is.na(officer_gender)) %>% nrow()
)
```

# Final cleanup and consistency checks
## Duplicate entries for DRBA PD
The "Delaware River and Bay Authority Police Department" is listed twice,
associated with two different counties, with with two entries for each officer.
```{r}
drbapd <- "Delaware River and Bay Authority Police Department"
agency %>% filter(agency_name==drbapd)
officer %>% filter(agency_name==drbapd) %>% 
  group_by(agency_county) %>%
  count()
```
We will keep just the Salem County records, and change the `agency_county` to `NA`.
```{r}
# In agency, filter out the non-Salem County DRBA agencies.
agency <- agency %>% filter(agency_name != drbapd |
                    (agency_name == drbapd &
                       agency_county == "Salem County"))
# Now replace Salem County with NA
agency$agency_county[agency$agency_name == drbapd] <- NA

# Now do the same for the officer table
officer <- officer %>% filter(agency_name != drbapd |
                              (agency_name == drbapd &
                                 agency_county == "Salem County"))
officer$agency_county[officer$agency_name == drbapd] <- NA
```
## Integrity check: uniqueness of primary key in `agency` table
Each combination of `year`, `agency_county` and `agency_name` should appear at most one time.
```{r}
stopifnot(
  0 ==
    agency %>%
    group_by(year, agency_county, agency_name) %>%
    count() %>%
    filter(n > 1) %>%
    nrow()
)
```
## Integrity check: foreign key in `officer` table
Every `year`, `agency_county` and `agency_name` appearing in `officer` should
also appear in `agency`.
```{r}
stopifnot(
  0 ==
    officer %>%
    select(year, agency_county, agency_name) %>%
    unique() %>%
    anti_join(agency, by = names(.)) %>%
    nrow()
)
```

## Integrity check: Every agency has at least one officer
For every `year`, `agency_county` and `agency_name` combination appearing in
the `agency` table, check that there is at least one entry in the `officer` table.
```{r}
stopifnot(
  0 ==
    agency %>%
    select(year, agency_county, agency_name) %>%
    anti_join(officer, by = names(.)) %>%
    nrow()
)
```

## Consistency with `njoaguof`.
Ideally, every agency in the `njoaguof` package would be reflected in the
the `agency` table.  However, the NJ OAG LEOD dataset excludes corrections 
officers. And even after we filter out corrections officers from the 
`njoaguof` dataset, there are still some missing agencies. 

Let's check that the only missing agencies are the various corrections 
departments, plus a short list of known exceptions.
```{r}
stopifnot(
  njoaguof::incident %>%
    select(agency_county, agency_name) %>%
    unique() %>%
    anti_join(agency, by = names(.)) %>%
    filter(!str_detect(agency_name, "Corrections")) %>%
    pull(agency_name) %in%
    c(
      "Toms River Township",
      "Hardyston Twp Police Dept",
      "Linden Police Department",
      "Freehold Boro PD",
      "Washington Twp PD",
      "College Of New Jersey",
      "State Parole Board"
    ) %>%
    all()
)
```
